---
title: "RNA-seq analysis with simpleSingleCell from Bioconductor"
author: "Thomas Walsh"
date: May 4, 2026
bibliography: references.bib
format:
  revealjs:
    slide-level: 2
    execute:
      echo: false
      eval: true
  gfm:
    output-file: README.md
    execute:
      echo: false
      eval: true
---

::: {.content-visible when-format="revealjs"}

## Overview
This workflow demonstrates a complete single-cell RNA-seq (scRNA-seq) analysis pipeline to Identify distinct brain cell populations (e.g., interneurons) from heterogeneous data.

* Process raw single-cell gene expression data
* Remove low-quality cells
* Normalize technical variation
* Identify biologically meaningful variation
* Cluster cells into distinct populations
* Detect marker genes to assign cell types

:::

::: {.content-visible when-format="revealjs"}

## Zeisel Mouse Brain Dataset

* ~3,000 single cells
* Tissue: Cortex & hippocampus
* Technology:
* Fluidigm C1 system
* Unique Molecular Identifiers (UMI)-based sequencing

Highly heterogeneous, which is ideal for clustering
Well-annotated for easy biological validation 

:::

::: {.content-visible when-format="revealjs"}

## Data Loading & Cleaning

* Loaded dataset into `SingleCellExperiment`
* Merged duplicate gene entries
* Fetch the Ensembl gene IDs

```{r 01_Setup}
# Dependencies needed
# Package lists
cran_packages <- c(
  "magrittr",
  "dplyr",
  "ggplot2",
  "pheatmap",
  "RColorBrewer",
  "generics",
  "matrixStats"
)

bioc_packages <- c(
  "BiocSingular",
  "scran",
  "org.Mm.eg.db",
  "AnnotationDbi",
  "scater",
  "scuttle",
  "scRNAseq",
  "SingleCellExperiment",
  "SummarizedExperiment",
  "Biobase",
  "GenomicRanges",
  "Seqinfo",
  "IRanges",
  "S4Vectors",
  "BiocGenerics",
  "MatrixGenerics",
  "BiocStyle",
  "rebook"
)

# Install any missing packages

# Install CRAN packages
cran_to_install <- cran_packages[!cran_packages %in% installed.packages()[, "Package"]]
if (length(cran_to_install) > 0) {
  install.packages(cran_to_install, repos = "https://cloud.r-project.org")
}

# Ensure BiocManager is available
if (!requireNamespace("BiocManager", quietly = TRUE)) {
  install.packages("BiocManager", repos = "https://cloud.r-project.org")
}

# Install Bioconductor packages
bioc_to_install <- bioc_packages[!bioc_packages %in% installed.packages()[, "Package"]]
if (length(bioc_to_install) > 0) {
  BiocManager::install(bioc_to_install, ask = FALSE, update = FALSE)
}

# Load all packages quietly
all_packages <- c(cran_packages, bioc_packages)

invisible(
  lapply(all_packages, function(pkg) {
    suppressPackageStartupMessages(
      library(pkg, character.only = TRUE)
    )
  })
)
```

```{r}
# Packages/calls
sce.zeisel <- ZeiselBrainData()

sce.zeisel <- aggregateAcrossFeatures(sce.zeisel, 
    id=sub("_loc[0-9]+$", "", rownames(sce.zeisel)))

# Fetch the Ensembl gene IDs, just in case they're needed them later.
rowData(sce.zeisel)$Ensembl <- mapIds(org.Mm.eg.db, 
    keys=rownames(sce.zeisel), keytype="SYMBOL", column="ENSEMBL")
```

:::

::: {.content-visible when-format="revealjs"}

## Quality Control
Remove low-quality cells
QC metrics:
  Library size
  Detected genes
  Mitochondrial %
  ERCC %

QC Results:
  Low features:         3 cells
  High ERCC:           65 cells
  High mitochondrial: 128 cells
  Total removed:      189

:::

```{r}
# base check
unfiltered <- sce.zeisel

stats <- perCellQCMetrics(sce.zeisel, subsets=list(
    Mt=rowData(sce.zeisel)$featureType=="mito"))
qc <- quickPerCellQC(stats, percent_subsets=c("altexps_ERCC_percent", 
    "subsets_Mt_percent"))
sce.zeisel <- sce.zeisel[,!qc$discard]

```

::: {.content-visible when-format="revealjs"}

```{r}
#| echo: false
#| fig-cap: "Figure 1: Distribution of each QC metric across cells in the Zeisel brain dataset. Each point represents a cell and is colored according to whether that cell was discarded."
#| fig-width: 8
#| fig-height: 6

colData(unfiltered) <- cbind(colData(unfiltered), stats)
unfiltered$discard <- qc$discard

p <- gridExtra::grid.arrange(
    plotColData(unfiltered, y="sum", colour_by="discard") +
        scale_y_log10() + ggtitle("Total count"),
    plotColData(unfiltered, y="detected", colour_by="discard") +
        scale_y_log10() + ggtitle("Detected features"),
    plotColData(unfiltered, y="altexps_ERCC_percent",
        colour_by="discard") + ggtitle("ERCC percent"),
    plotColData(unfiltered, y="subsets_Mt_percent",
        colour_by="discard") + ggtitle("Mito percent"),
    ncol=2
)

print(p)
```

:::

::: {.content-visible when-format="revealjs"}

```{r}
#| fig-cap: "Figure 2: Percentage of mitochondrial reads in each cell in the Zeisel brain dataset, compared to the total count (left) or the percentage of spike-in reads (right). Each point represents a cell and is colored according to whether that cell was discarded."
gridExtra::grid.arrange(
    plotColData(unfiltered, x="sum", y="subsets_Mt_percent",
        colour_by="discard") + scale_x_log10(),
    plotColData(unfiltered, x="altexps_ERCC_percent", y="subsets_Mt_percent",
        colour_by="discard"),
    ncol=2
)
```

:::

::: {.content-visible when-format="revealjs"}

## Normalization
Normalize for sequencing depth
Use deconvolution size factors

```{r}
# set seed
set.seed(1000)
clusters <- quickCluster(sce.zeisel)
sce.zeisel <- computeSumFactors(sce.zeisel, cluster=clusters) 
sce.zeisel <- logNormCounts(sce.zeisel)
```

```{r}
#| Fig-cap: "Figure 3: Relationship between the library size factors and the deconvolution size factors in the Zeisel brain dataset."
plot(librarySizeFactors(sce.zeisel), sizeFactors(sce.zeisel), pch=16,
    xlab="Library size factors", ylab="Deconvolution factors", log="xy")
```

:::

::: {.content-visible when-format="revealjs"}

## Variance Modeling
Identify highly variable genes
Separate biological vs technical variance

```{r}
dec.zeisel <- modelGeneVarWithSpikes(sce.zeisel, "ERCC")
top.hvgs <- getTopHVGs(dec.zeisel, prop = 0.1)
```

```{r}
#| fig-cap: "Figure 4: Per-gene variance as a function of the mean for the log-expression values in the Zeisel brain dataset. Each point represents a gene (black) with the mean-variance trend (blue) fitted to the spike-in transcripts (red)."
plot(dec.zeisel$mean, dec.zeisel$total, pch=16, cex=0.5,
    xlab="Mean of log-expression", ylab="Variance of log-expression")
curfit <- metadata(dec.zeisel)
points(curfit$mean, curfit$var, col="red", pch=16)
curve(curfit$trend(x), col='dodgerblue', add=TRUE, lwd=2)
```

:::

::: {.content-visible when-format="revealjs"}

## Dimensionality Reduction

* PCA for denoising
* t-SNE for visualization
* number of PCs retained by `denoisePCA()`.

```{r}
# set seed, run PCA and tSNE
set.seed(101011001)
sce.zeisel <- denoisePCA(sce.zeisel, technical=dec.zeisel, subset.row=top.hvgs)
sce.zeisel <- runTSNE(sce.zeisel, dimred="PCA")
```

```{r}
ncol(reducedDim(sce.zeisel, "PCA"))
```

:::

::: {.content-visible when-format="revealjs"}

## Clustering & t-SNE Visualization
* Each point = cell
* Color = cluster
* Graph-based clustering
* 14 clusters identified

```{r}
# cluster and labels
snn.gr <- buildSNNGraph(sce.zeisel, use.dimred="PCA")
colLabels(sce.zeisel) <- factor(igraph::cluster_walktrap(snn.gr)$membership)

# table for the cluster labels
table(colLabels(sce.zeisel))
```

```{r}
# t-SNE plot for labeled clusters
#| fig-cap: "Figure 5: Obligatory t-SNE plot of the Zeisel brain dataset, where each point represents a cell and is colored according to the assigned cluster."
plotTSNE(sce.zeisel, colour_by="label")
```

:::

::: {.content-visible when-format="revealjs"}

## Marker Detection
* Identify cluster-specific genes
* Compare each cluster vs others

Top markers for Cluster 1:
* Gad1
* Gad2
* Slc32a1

**Indicates interneurons**

```{r}
markers <- findMarkers(sce.zeisel, direction = "up")
marker.set <- markers[["1"]]
```

:::

::: {.content-visible when-format="revealjs"}

## Marker Heatmap
* Visualize expression across clusters

```{r}
#| fig-cap: "Figure 6: Heatmap of the log-expression of the top markers for cluster 1 compared to each other cluster. Cells are ordered by cluster and the color is scaled to the log-expression of each gene in each cell."
top.markers <- rownames(marker.set)[marker.set$Top <= 10]
plotHeatmap(sce.zeisel, features=top.markers, order_columns_by="label")
```

:::

::: {.content-visible when-format="revealjs"}

## LogFC Heatmap
* Compare marker strength across clusters

```{r}
# heatmap, alt - to see log-fold changes
#| fig-cap: "Figure 7: Heatmap of the log-fold changes of the top markers for cluster 1 compared to each other cluster."
logFCs <- getMarkerEffects(marker.set[1:50,])
pheatmap(logFCs, breaks=seq(-5, 5, length.out=101))
```

:::

::: {.content-visible when-format="revealjs"}

## Summary
* Processed scRNA-seq dataset
* Identified 14 cell clusters
* Detected biologically meaningful markers
* Interneurons identified in Cluster 1

:::

::: {.content-visible when-format="revealjs"}

## References

* Pollen, A. A., T. J. Nowakowski, J. Shuga, X. Wang, A. A. Leyrat, J. H. Lui, N. Li, et al. 2014. “Low-coverage single-cell mRNA sequencing reveals cellular heterogeneity and activated signaling pathways in developing cerebral cortex.” Nat. Biotechnol. 32 (10): 1053–8.

* Zeisel, A., A. B. Munoz-Manchado, S. Codeluppi, P. Lonnerberg, G. La Manno, A. Jureus, S. Marques, et al. 2015. “Brain structure. Cell types in the mouse cortex and hippocampus revealed by single-cell RNA-seq.” Science 347 (6226): 1138–42.

* Zeng, H., E. H. Shen, J. G. Hohmann, S. W. Oh, A. Bernard, J. J. Royall, K. J. Glattfelder, et al. 2012. “Large-scale cellular-resolution gene profiling in human neocortex reveals species-specific molecular signatures.” Cell 149 (2): 483–96.

```{r}
#| echo: false
#| eval: false

# Reference information for the bibliography obtained using citation() 

# core Analysis & Infrastructure
citation("SingleCellExperiment")
citation("SummarizedExperiment")
citation("GenomicRanges")
citation("IRanges")
citation("S4Vectors")
citation("BiocGenerics")

# Single-Cell Specific Tools
citation("scran")
citation("scater")
citation("scuttle")
citation("scRNAseq")
citation("BiocSingular")

# Annotation & Data
citation("org.Mm.eg.db")
citation("AnnotationDbi")
citation("Biobase")
citation("Seqinfo")

# Utility & Visualization
citation("rebook")
citation("BiocStyle")
citation("MatrixGenerics")
citation("matrixStats")
citation("generics")
citation("RColorBrewer")
citation("pheatmap")
citation("ggplot2")
citation("dplyr")
citation("magrittr")
```

:::